---
title: "getweatherdata"
author: "Aviel Fradkine"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
library(tidyverse)
library(readxl)
library(ggthemes)
library(haven)
library(huxtable)
library(knitr)
library(weathercan)
library(sp)
library(lubridate)
knitr::opts_chunk$set(
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  comment = "",
  echo = TRUE
)
```

## weatherdata file part

Manipulate weather station data to get what we want
```{r}
station_inventory <- read_csv("station_inventory.csv", skip = 3)

# filter for stations where there is data for first and last years 1991 - 2021
station_inventory_1991to2021 <- station_inventory %>%
        filter(`DLY First Year` < 1992 & `DLY Last Year` > 2020)

# filter by latitude lower than 55 (to exclude more northern territories)
station_inventory_1991to2021_filtered <- station_inventory_1991to2021 %>%
        filter(`Latitude (Decimal Degrees)` < 55)

station_lists_1991to2021_filtered <- station_inventory_1991to2021 %>%
        select(cols = c(`Name`, `Province`, `Station ID`, `Climate ID`))

colnames(station_lists_1991to2021_filtered) <- c('Name', 'Province', 'Station ID', 'Climate ID')

write_csv(station_inventory_1991to2021_filtered, "C:\\Users\\Aviel\\Desktop\\Econ468project\\dailystations1991to2021filtered.csv")
write_csv(station_lists_1991to2021_filtered, "C:\\Users\\Aviel\\Desktop\\Econ468project\\stationnames1991to2021filtered.csv")
```

## Nov 9 2022

Exploring the stations

```{r}
daily <- read_csv("dailystations1991to2021filtered.csv")

daily_alberta <- daily[daily$Province == "ALBERTA", ]
daily_manitoba <- daily[daily$Province == "MANITOBA", ]
daily_saskatchewan <- daily[daily$Province == "SASKATCHEWAN", ]
daily_ontario <- daily[daily$Province == "ONTARIO", ]
daily_quebec <- daily[daily$Province == "QUEBEC", ]

all_stations_filtered <- rbind(daily_alberta, daily_manitoba, daily_saskatchewan, daily_ontario, daily_quebec)

rm(list = setdiff(ls(), "all_stations_filtered")) # keeps only all stations filtered object
```

Combine all the stations into one frame
```{r}
setwd("C:/Users/Aviel/Desktop/Econ468project/all_filtered_stations_1991to2021/")
a <- list.files("C:/Users/Aviel/Desktop/Econ468project/all_filtered_stations_1991to2021/")

length(a) 

# read in all the station data 
for (i in 1:length(a)){
        setwd("C:/Users/Aviel/Desktop/Econ468project/all_filtered_stations_1991to2021/")
        assign(print(a[i]), read_csv(a[i]))
}

all_stations_data <- get(a[1])

for (i in 2:length(a)){
        all_stations_data <- rbind(all_stations_data, get(a[i]))
        print(i)
}

# write.csv(all_stations_data, "C:/Users/Aviel/Desktop/Econ468project/all_stations_data.csv") 

rm(list = setdiff(ls(), "all_stations_copy")) # remove all the other objects
```

## Data exploring

```{r}
all_stations_data <- read.csv("C:/Users/Aviel/Desktop/Econ468project/all_stations_data.csv")

all_stations_copy <- all_stations_data %>%
        select(cols = -c(`...1`)) # remove unnecessary column

# create station name vector
station_IDs <- unique(all_stations_copy$station_id)

filtered_stations <- read_csv("C:/Users/Aviel/Desktop/Econ468project/stationnames1991to2021filtered.csv")

# gives us 214 rows with each station being only the ones we use
filtered_stations <- filtered_stations %>%
        filter(`Station ID` %in% station_IDs)

# count number of NAs for each stations mean temp data
n_NAs_by_station <- all_stations_copy %>%
        group_by(`station_id`) %>%
        summarize(number_of_NAs = sum(is.na(`mean_temp`)))

# create a third column with percentage NA
# create a fourth column with province code
n_NAs_by_station <- n_NAs_by_station %>%
        mutate(percent_NAs = `number_of_NAs`/nrow(n_NAs_by_station)) %>%
        mutate(prov = case_when(
                (station_id %in% c(1824:2833)) ~ "AB",
                (station_id %in% c(2854:3319)) ~ "SK",
                (station_id %in% c(3462:3858)) ~ "MB",
                (station_id %in% c(3920:5171)) ~ "ON",
                (station_id %in% c(5216:7000))~ "QC"
        ))

# fix missing case with Manitoba 
n_NAs_by_station[51:53, 4] <- c("MB")

# fix missing case with Quebec
n_NAs_by_station[114, 4] <- c("QC")

# sort by number of NA's
n_NAs_by_station <- n_NAs_by_station %>%
        group_by(`prov`) %>%
        arrange(`percent_NAs`, .by_group = TRUE)

stations_alberta <- n_NAs_by_station %>%
        filter(`prov` == c("AB"))
stations_manitoba <- n_NAs_by_station %>%
        filter(`prov` == c("MB"))
stations_sask <- n_NAs_by_station %>%
        filter(`prov` == c("SK"))
stations_ontario <- n_NAs_by_station %>%
        filter(`prov` == c("ON"))
stations_quebec <- n_NAs_by_station %>%
        filter(`prov` == c("QC"))

# ASSEMBLE SAMPLE
# 17 stations from alberta 
sample_alberta <- stations_alberta %>%
        slice(c(1:17))
# 28 stations from saskatchewan
sample_saskatchewan <- stations_sask %>%
        slice(c(1:28))
# 8 stations from manitoba
sample_manitoba <- stations_manitoba %>%
        slice(c(1:8))
# 11 stations from ontario 
sample_ontario <- stations_ontario %>%
        slice(c(1:11))
# 3 stations from quebec 
sample_quebec <- stations_quebec %>%
        slice(c(1:3))

test_sample_station_NAs <- rbind(sample_alberta, sample_saskatchewan, sample_manitoba, sample_ontario, sample_quebec)

test_sample_station_IDs <- test_sample_station_NAs$station_id
        
## Take all the data for those stations 

all_stations_sample <- all_stations_copy %>%
        filter(`station_id` %in% test_sample_station_IDs)
```

Now, using the data from the filtered set, make the temperature index
```{r}
all_stations_sample_weather <- all_stations_sample %>%
        select(`station_name`, `station_id`, `lat`, `lon`, `elev`,
               `date`, `year`, `month`, `day`, `max_temp`, `mean_temp`, `min_temp`, `total_precip`)

# create a third column which has the month year combo as a date 
all_stations_sample_weather$month_year <- format(as.Date(all_stations_sample_weather$date), "%Y-%m")

# order by date
all_stations_sample_weather <- all_stations_sample_weather %>%
        group_by(`date`, `month_year`) %>%
        arrange(date, .group_by = T)

# write.csv(all_stations_sample_weather, "C:/Users/Aviel/Desktop/Econ468project/sample_weather.csv")
```

[NOV 14] 
IMPORTANT: To resume work from the ordered data set for all the stations we are using use data in "sample_weather.csv"

```{r}
## create a weather index with mean temp by day for min, max, and mean temp
# each day and the average of mean temp across all stations
avg_mean_temp_by_day <- all_stations_sample_weather %>%
        group_by(`month_year`, `date`) %>%
        summarize(avg_mean_temp = mean(mean_temp, na.rm = TRUE))
# each day and the average of min temp across all stations
avg_min_temp_by_day <- all_stations_sample_weather %>%
        group_by(`month_year`, `date`) %>%
        summarize(avg_min_temp = mean(min_temp, na.rm = TRUE))

# each day and the average of max temp across all stations
avg_max_temp_by_day <- all_stations_sample_weather %>%
        group_by(`month_year`, `date`) %>%
        summarize(avg_max_temp = mean(max_temp, na.rm = TRUE))

# create a factor for the mean temp by day for month
avg_mean_temp_by_month <- avg_mean_temp_by_day %>%
        group_by(month_year) %>%
        summarize(month_mean_temp = mean(avg_mean_temp)) %>%
        arrange(month_year)

# write.csv(avg_mean_temp_by_month, "C:/Users/Aviel/Desktop/Econ468project/regression_inputs/monthly_mean_temp.csv")

```

Constructing the index for max temp
```{r}
daily_max_30 <- all_stations_sample_weather %>%
        summarize(n_days_max_over_30 = sum(max_temp >= 30, na.rm = TRUE))

daily_max_25 <- all_stations_sample_weather %>%
        summarize(n_days_max_over_25 = sum(max_temp >= 25, na.rm = TRUE))

# summarize over months the number of stations in month per days which had over 30
monthly_max_30 <- daily_max_30 %>%
        group_by(`month_year`) %>%
        summarize(total_station_max_days_over_30 = sum(n_days_max_over_30))
# write.csv(monthly_max_30, "C:/Users/Aviel/Desktop/Econ468project/regression_inputs/monthly_max_30.csv")

# summarize over months the number of stations in month per days which had over 25
monthly_max_25 <- daily_max_25 %>%
        group_by(`month_year`) %>%
        summarize(total_station_max_days_over_25 = sum(n_days_max_over_25))
# write.csv(monthly_max_25, "C:/Users/Aviel/Desktop/Econ468project/regression_inputs/monthly_max_25.csv")
```

Now let's try to construct the index for mean
```{r}
#data set
daily_mean_30 <- all_stations_sample_weather %>%
        summarize(n_days_over_30 = sum(mean_temp >= 30, na.rm = TRUE))

daily_mean_25 <- all_stations_sample_weather %>%
        summarize(n_days_over_25 = sum(mean_temp >= 25, na.rm = TRUE))

# summarize over months the number of stations in month per days which had over 30
monthly_mean_30 <- daily_mean_30 %>%
        group_by(`month_year`) %>%
        summarize(total_station_days_over_30 = sum(n_days_over_30))
# write.csv(monthly_mean_30, "C:/Users/Aviel/Desktop/Econ468project/regression_inputs/monthly_over_30.csv")

# summarize over months the number of stations in month per days which had over 25
monthly_mean_25 <- daily_mean_25 %>%
        group_by(`month_year`) %>%
        summarize(total_station_days_over_25 = sum(n_days_over_25))
# write.csv(monthly_mean_25, "C:/Users/Aviel/Desktop/Econ468project/regression_inputs/monthly_over_25.csv")
```

Precipitation data
```{r}
precipitation_monthly <- all_stations_sample_weather %>%
        group_by(`month_year`) %>%
        summarize(monthly_precip = sum(total_precip, .by_group = T, na.rm = T)) %>%
        mutate(avg_precip = monthly_precip/length(unique(all_stations_sample_weather$station_name)))
# write.csv(precipitation_monthly, "C:/Users/Aviel/Desktop/Econ468project/regression_inputs/monthly_precipitation.csv```

```
